---
title: "Natural Language Processing"
output: html_document
---

```{r, message=FALSE}
library(dplyr)
library(ggplot2)
library(readr)
library(stringi)
library(tidyr)
library(magrittr)

theme_set(theme_minimal())
```

This RMarkdown script is a revised and condensed version of ninth chapter from
the book "Humanities Data in R". There have been many changes over the past 5
year in regards to NLP technology and this revision uses our new R package (cleanNLP)
in place of the older and unsupported coreNLP. The book itself is still useful
as it contains additional information about each of the NLP tasks. We suggest pairing
the exposition in the book but to run the R code from this file instead.

# Tokenization & Sentence Splitting

Consider (an English translation of) the opening lines to Albert Camus's
*l'Étranger*. Let's represent this as a length one character vector in R.

```{r}
camus <- "Mother died today. Or, maybe, yesterday; I can't be sure."
```

A practical, and seemingly simple, first step in processing this string
is to split it into a longer character vector where each element
contains a single word. But what exactly is meant by a word? Are
punctuation marks separate words? What about contractions, hyphens, or
compound nouns such as 'New York City'? If we split this string apart
using just the presence of spaces the result seems approximately
reasonable, but less than perfect.

```{r}
stri_split(camus, fixed=" ")
```

The process of splitting text into meaningful elements is called tokenization.
For an English text the difficult task is catching the myriad of rules and
exceptions. Rather than re-creating and re-implementing these conditions, it
is better to use a well-tested library to tokenize our string.

Throughout this chapter will be make use of an R package we have
developed called **cleanNLP**. The package provides various backends that
use different libraries to perform various NLP tasks, but then transforms
their output into a common format. To start, we will use the backend
because it is reasonably powerful but also easy to install (in fact, you
should not need to do anything other than install cleanNLP from CRAN). To
start, we load the package and call to the function to indicate that we
want to use this particular backend.

```{r}
library(cleanNLP)
cnlp_init_udpipe()
```

The first time you run this function, R will automatically download the
language model from the internet. This may take a minute or two
depending on your internet connection; the file will be cached for
future use.

In order to process a string of text using we pass a text through the
function . We will explore the various parts of the output throughout
this chapter.

```{r}
annotation <- cnlp_annotate(camus)
names(annotation)
```

The two elements of the annotation give information about the detected
tokens and the input documents, respectively. Most of the useful
information is in the tokens table, such as an improved version of
tokenization as seen here:

```{r}
token <- annotation$token
token$token
```

Notice that the punctuation symbols have been assigned to their own
elements and the contraction 'can't' has been split into two words.
(Note: In the next step, 'ca' and 'n't' will be addressed.) However, the
input text has not been modified with the exception of removing spaces.
In addition to the process of tokenizing the input, the annotation also
calculated how to split the input into sentences. To see the assignment
of sentences, pull the sentence element out of the tokenization.

```{r}
token$sid
```

The algorithm has assigned the first four tokens to the first sentence,
and the remainder to a second sentence.

Can we learn anything useful directly from the output of this process of
tokenization? In our short $16$ lemma example probably not, but consider
applying the algorithm to a longer sample of text. We will now try to
use a longer text, here the first published Sherlock Holmes short story,
"A Scandal in Bohemia".

```{r}
text <- readLines("data/holmes/01_a_scandal_in_bohemia.txt")
anno <- cnlp_annotate(paste(text, collapse="\n"))
token <- anno$token
dim(token)
```

The output object is of exactly the same structure as our short example.
In this case 10587 tokens have been processed. We can use the
tokenization and sentence splitting to determine the length of every
sentence in the text.

```{r}
token %>%
  group_by(doc_id, sid) %>%
  summarize(sent_len = length(tid)) %>%
  ggplot(aes(sent_len)) +
    geom_histogram(bins=30, color="black", fill="white")
```

The distribution shows a sharp peak of sentences less then 10 tokens long;
this is fairly short, particularly when considering that punctuation is included in this
count, and likely a product of the heavy use of dialogue in the text. A
very small set of sentences are 60 or more tokens long. A close analysis of
the original text reveals most of these to be part of a long deductive
speech given by Sherlock Holmes towards the end of the story.

# Lemmatization & Part of Speech Tagging

While tokenizing simply splits the raw character input into groups,
lemmatization goes further by converting each token into a representative
lemma. For example, 'go' is the English lemma for words such as 'gone',
'going', and 'went'. Nouns have their own process of lemmatization, such as
converting all words into their singular form; for example 'dogs' becomes
'dog' and 'mice' becomes 'mouse'.

Notice that lemmatization changes depending on the part of speech.
Therefore, much like tokenization and sentence splitting, the task of
tagging tokens with parts of speech and lemmatization is often
accomplished in tandem. To demonstrate how both work in the package, we
again turn to our annotation of 'A Scandal in Bohemia'. Pulling out the
second sentence (and obscuring some of the columns so as not to overwhelm
ourselves with the rest of the output) reveals the following data frame.

```{r}
select(token, doc_id, sid, tid, token, lemma, xpos)
```

The lemmatization process should seem straightforward. The verb 'heard'
is now represented by the infinitive 'hear', and the pronouns 'him' and
'her' are changed to their nominative forms 'he' and 'she'. Otherwise,
the words remain unchanged in their lemma form. The part of speech
codes, on the other hand, may at first seem confusing; for example there
are three different codes for the verbs 'have', 'heard', and 'mention'.

The part of speech codes come from the Penn Treebank Project, and contain
many more categories compared to those typically taught in primary school
grammar courses. For example **VBN** is the past participle form of a verb,
whereas **VB** is the base form of a verb. A table from our annotation shows
the entire set of possibilities.

```{r}
token %>%
  count(xpos, sort = TRUE)
```

We see that **JJS**, superlative adjectives, are relatively uncommon,
and **NN** and **IN** tags, prepositions or subordinating conjunction, occur
quite frequently. For a complete description of these tags see the technical
report and justification from the Penn Treebank Project.

The extended set of parts of speech are quite useful, but many times a
smaller set of more familiar options can better serve a particular
purpose. Unviversal tagsets have been created as language-agnostic part
of speech classifiers. A mapping from Penn Treebank
codes into this smaller tag-set is provided as a column in the output. A table
using universal tagsets reveals a smaller and more familiar list of parts of
speech.

```{r}
token %>%
  count(upos, sort = TRUE)
```

We can see how these map to the more grainular part of speech codes as follows:

```{r}
token %>%
  count(upos, xpos) %>%
  filter(upos %in% c("NOUN", "PROPN", "VERB"))
```

Nouns correspond to plural nouns (**NNS**) and singular (**NN**) nouns.
There are also two plural proper noun codes (**NNP** and **NNPS**). The
verb subtypes refer to various broad categories of verb conjugations.

With these universal part of speech codes, we can run some basic
analysis on the sentences in our text. First, we count the number of
nouns, pronouns, adjectives and verbs in each sentence using the
function.

```{r}
pos_df <- token %>%
  group_by(doc_id, sid, upos) %>%
  summarize(n = n()) %>%
  pivot_wider(names_from="upos", values_from=n, values_fill = list(n=0))
pos_df
```

An interesting stylistic question would be to see the within-sentence
distribution of nouns (including pronouns), verbs, and adjectives. Due
to the discrete nature of count data, a simple scatter plot will not
work well. Instead we take advantage of over-plotting and color opacity
to produce an interpretable visual description of this distribution.

```{r}
pos_df %>%
  ggplot(aes(NOUN + PRON, VERB)) +
    geom_point(size=2, alpha=0.1, color="blue")
```

```{r}
pos_df %>%
  ggplot(aes(NOUN + PRON, ADJ)) +
    geom_point(size=2, alpha=0.1, color="blue")
```

We see that verbs and nouns are well correlated
and roughly appear in equal numbers for short sentences, whereas longer
sentences tend to increase in terms of nouns faster than the number of
verbs. Adjectives do not follow such a smooth relationship with nouns;
regardless of the number of nouns they rarely occur more than 2 times in
a sentence, despite occurring once in almost half of sentences with only
one noun. One cause for both of these effects comes from the fact that
multi-word nouns get double counted. For instance the name 'Dr. John H.
Watson', accounts for $4$ noun counts, but would not generally require
more adjectives or verbs than when 'Watson' is used alone.

Part of speech tagging also has the benefit of isolating function words,
those with primarily grammatical usage such as prepositions and conjunctions, from
those words with lexical meaning. Along with the use of lemmatization to
collapse various word forms together, we can now gain potentially
important contextual information from a text by identifying the most
frequently used lemmas from a particular part of speech. The top $25$
noun lemmas from our sample text for example identify some of the key
characters and objects of interest in the story.

```{r}
token %>%
  filter(upos %in% c("NOUN", "PROPN")) %>%
  count(lemma) %>%
  arrange(desc(n))
```

The third most common word, 'photograph', comes from the main objective
presented in the story: the recovery of a scandalous photograph.
References to the main characters are also present: 'Sherlock Holmes' of
course, 'Majesty' being the royal client wishing to recover the
photograph, and 'Irene Adler' the subject of the photograph in question.

Knowing beforehand the text in question, we were able to extract the
character names from a table of the most common nouns. Identifying facts
we already know however is not particularly useful. We can get a bit closer
by looking only at proper nouns:

```{r}
token %>%
  filter(upos %in% c("PROPN")) %>%
  count(lemma) %>%
  arrange(desc(n))
```

The reduced list includes an increased set of characters, while removing
many of the non-name nouns from the list. The results are still lacking
as the titles and first and last names are not linked together. Some
non-names are also present, such as the proper place names 'Europe' and
'London'. In order to resolve these issues we need to discern
relationships between pairs of words, rather than working with lemmas
individually.

# Dependencies

To this point, we have primarily worked with individual words, tokens,
and lemmas. We now approach the subject of sentence parsing, where the words
within a sentence are assigned a complete linguistic structure linking
together all of the individual parts. The result of this, known as a
parse tree, has a nice graphical structure. Looking at the additional
columns in the token table, here are the dependencies that were detected
by .

```{r}
token %>%
  filter(sid == 5)
```

As with the part of speech codes, the dependency type codes can be
difficult to interpret without a code book. The one that we
will make explicit use of here is **nsubj**, which identifies 'a noun phrase,
which is syntactic subject of a clause'. The latter typically relates a noun
and its verb, but may link a noun to another noun (as above) or adjective in
the presence of a copular verb.

The output from cleanNLP gives a compact representation of the dependency
structure. It will be useful to include more information in the table to
tie together each token with its source dependency. This is achieved
through joining the tokens table with itself.

```{r}
dep <- token %>%
  left_join(
    select(token, doc_id, sid, tid_source=tid, token, lemma, upos, xpos),
    suffix=c("", "_dep"),
    by=c("doc_id", "sid", "tid_source")
  )
dep
```

Now, for example, we can see that the token 'I' is the subject of the
verb 'here'. For an example of how dependency information can be used to
understand a textual source, consider identifying the most frequently used
verbs that take the action from the pronoun 'I' used as the subject of a
sentence.

```{r}
dep %>%
  filter(token == "I", relation == "nsubj") %>%
  count(lemma_dep) %>%
  arrange(desc(n))
```

The analysis here is slightly difficult because 'I' may represent the
narrator or could be present in a quotation.

# Named Entity Recognition

The task of automatically detecting and classifying elements of a text
into broad semantic categories is known as named entity recognition (NER).
To get NERs from cleanNLP, we need to use a different backend. Here, we
will use the **spacy** backend. It is faster and more accurate, but does
take a bit more work to install (see https://github.com/statsmaths/cleanNLP
for more details). Here we initalize the backend and run it over our text:

```{r}
cnlp_init_spacy()
text <- readLines("data/holmes/01_a_scandal_in_bohemia.txt")
anno <- cnlp_annotate(paste(text, collapse="\n"))
token <- anno$token
ner <- anno$entity

ner
```

Taking our sample text, we see the following $11$ categories picked up
by the algorithm ('O' is used for a non-hit).

```{r}
ner %>%
  count(entity_type, sort = TRUE)
```

Which kinds of tags were identified as locations (either LOC or GPE)?
Thankfully, many of the false positives in our character set are picked up
by the location tag. See for instance New Jersey and Baker Street.

```{r}
ner %>%
  filter(entity_type %in% c("LOC", "GPE")) %>%
  count(entity, sort = TRUE)
```

These can be helpful in determining the location or topic of a given
story.

The named entity tag that will help the most with our goal of
identifying the main characters in our short story, unsurprisingly, is
the one indicating persons.

```{r}
ner %>%
  filter(entity_type %in% c("PERSON")) %>%
  count(entity, sort = TRUE)
```

The resulting list of characters is a significant improvement over our
last attempt, with many false positives removed and longer names
reconstructed.

We could also clean up the duplicated names, such as 'Holmes'. This
often occurs when a character is mentioned at some point with only part
of their name. We do this by cycling through the list of names and
replacing the text of any one that is a strict subset of another.

```{r}
characters <- ner %>%
  filter(entity_type %in% c("PERSON")) %>%
  count(entity, sort = TRUE) %>%
  select(-n)

characters$full_name <- ""
for (j in seq_len(nrow(characters))) {
   these <- which(stri_detect(characters$entity, fixed=characters$entity[j]))
   these <- these[which.max(stri_length(characters$entity[these]))]
   characters$full_name[j] <- characters$entity[these]
}

characters$full_name <- stri_replace_all(characters$full_name, "", fixed="'")
characters %>%
  count(full_name, sort=TRUE)
```

# Case Study: Sherlock Holmes Main Characters

We have now built an algorithm for extracting character names using the
cleanNLP library and verified that the results were reasonable on a
single short story. Here we process the entire set of 56 short stories
using **cleanNLP**.

```{r}
cnlp_init_spacy()

text_names <- dir("data/holmes/", full.names=TRUE)
texts <- sapply(text_names, function(v) paste(readLines(v), collapse="\n"), USE.NAMES = FALSE)
anno <- cnlp_annotate(texts)
token <- anno$token
ner <- anno$entity
```

Now, we can identify character names using NER.

```{r}
longest_word <- function(v) { v[which.max(stri_length(v))] }

ner_characters <- ner %>%
  filter(entity_type == "PERSON") %>%
  left_join(select(token, doc_id, sid, tid, token), by=c("doc_id"="doc_id", "sid"="sid", "tid_end"="tid")) %>%
  group_by(doc_id, token) %>%
  mutate(char_name = longest_word(entity)) %>%
  ungroup()

ner_characters %>%
  count(doc_id, char_name, sort=TRUE)
```

And finally, can detect the most frequently mentioned character in each text that is neither
Sherlock Holmes nor John Watson.

```{r}
top_characters <- ner_characters %>%
  filter(!(token %in% c("Holmes", "Watson"))) %>%
  group_by(doc_id, char_name) %>%
  summarize(n = n()) %>%
  top_n(n, n=1) %>%
  select(-n)

top_characters$char_name
```

The results are encouraging and reveal an interesting list of characters
(none of which are repeated). Pulling out any particular story tends to
reveal a good match between our algorithm and a synopsis of the story.
Take for example number 23 'The Final Problem' and its introduction of the
criminal mastermind Professor Moriarty.

Other than just looking at this list of characters, there is substantial
analysis that can also be done with our parsed data. We can compute where
in each story a character is mentioned. The entire set of these can be
visualized for a representation of when and how often each character is referenced.

```{r, message=FALSE, fig.height=6}
ner_characters %>%
  semi_join(top_characters) %>%
  left_join(summarize(group_by(token, doc_id), total_sid = max(sid))) %>%
  mutate(percent = sid / total_sid) %>%
  ggplot(aes(percent, basename(text_names[doc_id]))) +
    geom_point() +
    xlab("Percentage of Story") + ylab("")
```

Notice, for instance, that Irene Adler has relatively few mentions in the the
first story compared to other characters in later installments.

# Other Languages

The cleanNLP package has support for parsing text in languages other
than English. To use these, simply specify the correct model when
running the function `cnlp_init_udpipe` or `cnlp_init_spacy`. For
example, here are the first two sentences of Albert Camus's L'Étranger
in French:

```{r}
camus_fr <- "Aujourd'hui, maman est morte. Ou peut-être hier, je ne sais pas. J'ai reçu un télégramme de l'asile."
```

Here is what happens when we use the English model:

```{r}
cnlp_init_udpipe()
anno <- cnlp_annotate(camus_fr)$token
anno$token
```

Not terrible, but it incorrectly splits the single tokens "Aujourd'hui" and "peut-être"
into two parts but does not split the two tokens "J'ai". The other parts of the NLP
pipeline are even more effected. Look, for example at the lemmas and part of speech
codes, which are almost all incorrect:

```{r}
anno
```

If we instead load the French model, these issues are fixed:

```{r, message=FALSE}
cnlp_init_udpipe("french")
anno <- cnlp_annotate(camus_fr)$token
anno
```

For example, "J'ai" is not split into two tokens which have the lemmas "Je" (I) and
"avoid" (to have). In general, it is important to use the correct language model when
working with textual corpora.
